Davos: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-07-16 paper/main.tex
fixed corresponding author notation issue

\title{Davos: a Python package ``smuggler'' for constructing
  lightweight reproducible notebooks}
\author{Paxton C. Fitzpatrick}
\author{Jeremy R. Manning\textsuperscript{\textsubscript{\scriptsize *}}}
\ead{Jeremy.R.Manning@Dartmouth.edu}
\cortext[cor]{Corresponding author}
\address{Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755}

@@

Since its initial release, Davos has found use in a variety of
applications. In addition to managing computing environments for multiple prior
and ongoing research studies~\citep{MannEtal23, OwenMann23, ZimaEtal23},
Davos is being used by both students and instructors in programming
and methods courses such as Storytelling with Data~\cite{Mann21d} (an open
course on data science, visualization, and communication) and Laboratory in
Psychological Science~\cite{Mann22} (an open course on experimental and
statistical methods for psychology research) to simplify distributing lessons
and submitting assignments, as well as in online demos such as
\texttt{abstract2paper}~\cite{Mann21e} (an example application of
GPT-Neo~\cite{GaoEtal20, BlacEtal21}) to share ready-to-run code that installs
dependencies automatically.

@@

While Davos enables developers to conveniently specify all project
dependencies, there are some edge cases and limitations that are worth
considering. Many Python packages include (in their \texttt{setup} options)
additional dependencies that often carry their own version specifications.
Although Davos will check that the correct version of the requested
top-level package is installed and imported into the workspace, the version
numbers of any \textit{dependencies} of the request package are \textit{not}
checked. In principle, this could lead to unexpected behavior, for example if a
given package's dependencies (or dependencies of those dependencies, etc.) were
left under-specified. A developer could mitigate this by explicitly smuggling
exact version numbers of \textit{every} project dependency (e.g., obtained via
\texttt{pip freeze}). However, for projects where the versions of dependencies
of \texttt{smuggle}d packages also need to be precisely controlled, a lockfile
or a \texttt{requirements.txt} file produced by \texttt{pip freeze} (i.e.,
explicitly specifying \textit{all} package's version numbers) may provide a
more comprehensive alternative to Davos.

Reproducibility is not solely about dependency management. In addition to
ensuring that project dependencies are satisfied, the user running a given
notebook must also execute the code in the indicated order. For example, the
cells in a notebook may be manually run out of order. If different cells in a
Davos-enhanced notebook made use of different versions of the same
package, this could result in \textit{more} confusion or \textit{greater}
replication failure rates relative to standard Jupyter notebooks. Therefore an
important consideration when using Davos is that it is perhaps even more
important to execute notebook cells in order than would be the case in the
standard (non-Davos) setup.  We suggest that developers who wish to use
Davos include notes regarding which cells must be executed in sequence.

As of this writing, Davos can install packages using \texttt{pip}, but
not other standard Python package management systems such as \texttt{conda}.
Therefore packages that are not installable via \texttt{pip} are currently
unsupported by Davos.  We anticipate adding support for other
package management systems, including \texttt{conda} in a future release.

>>>>

\title{Davos: a Python package ``smuggler'' for constructing
  lightweight reproducible notebooks}
\author{Paxton C. Fitzpatrick}
\author{Jeremy R. Manning\corref{cor}}
\ead{Jeremy.R.Manning@Dartmouth.edu}
\cortext[cor]{Corresponding author}
\address{Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755}

@@

Since its initial release, Davos has found use in a variety of applications. In
addition to managing computing environments for multiple prior and ongoing
research studies~\citep{MannEtal23, OwenMann23, ZimaEtal23}, Davos is being
used by both students and instructors in programming and methods courses such
as Storytelling with Data~\cite{Mann21d} (an open course on data science,
visualization, and communication) and Laboratory in Psychological
Science~\cite{Mann22} (an open course on experimental and statistical methods
for psychology research) to simplify distributing lessons and submitting
assignments, as well as in online demos such as
\texttt{abstract2paper}~\cite{Mann21e} (an example application of
GPT-Neo~\cite{GaoEtal20, BlacEtal21}) to share ready-to-run code that installs
dependencies automatically. The 2023 offering of Neuromatch
Academy~\cite{Neuromatch} also included an ``experimental'' module that uses
Davos to manage dependencies related to a large language model-based
tutor~\cite{Chatify}.

@@

While Davos enables developers to conveniently specify all project
dependencies, there are some edge cases and limitations that are worth
considering. First, reproducibility is not solely about dependency management.
In addition to ensuring that project dependencies are satisfied, the user
running a given notebook must also execute the code in the indicated order. For
example, the cells in a notebook may be manually run out of order. If different
cells in a Davos-enhanced notebook made use of different versions of the same
package, this could result in \textit{more} confusion or \textit{greater}
replication failure rates relative to standard Jupyter notebooks. Therefore an
important consideration when using Davos is that it is perhaps even more
important to execute notebook cells in order than would be the case in the
standard (non-Davos) setup. One approach to mitigating this risk for notebooks
that use several versions of the same library would be to include
\texttt{smuggle} statements in the same cell(s) where the library was called. A
second approach would be for developers who wish to use Davos to include notes
regarding which cells must be executed in sequence. This is already a common
practice for notebooks that include system calls to install required packages
and dependencies, and the same approach would work well for Davos-enhanced
notebooks as well.

A second limitation of Davos relates to how packages are installed and managed.
As of this writing, Davos can install packages using \texttt{pip}, but not
other standard Python package management systems such as \texttt{conda}.
Therefore packages that are not installable via \texttt{pip} are currently
unsupported by Davos. We anticipate adding support for other package management
systems, including \texttt{conda}, in a future release.

A third limitation of Davos is that it cannot be used to manage projects that
depend on non-Python software. For example, system software or libraries from
other languages (e.g., in a mixed Python and R notebook), cannot be
\texttt{smuggle}d by Davos. A notebook that utilizes or depends on non-Python
software would therefore need to use existing non-Davos approaches to managing
those requirements.
